---
title:  "TMB Basics"
subtitle: "TMB Training Session I"
author: "Andrea Havron<br>NOAA Fisheries, OST"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css", "slides-style.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


<style type="text/css">

code.cpp{
  font-size: 14px;
}
code.r{
  font-size: 14px;
}


</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()
```
<!-- modified from Cole Monnahan's TMB training series:  https://github.com/colemonnahan/tmb_workshop-->


---

# ML with TMB:
<br><br>
1. Write C++ model to specify the negative log-likelihood for a set of parameters and data
2. Compile the model and link to R
3. Construct the computational graph (tape)
3. Pass the following to a R minimizer:
 * objective function: specified by user in C++
 * initial parameters and data: specified by user in R
 * gradient functions: calculated by TMB based on objective function
4. At each step of minimization:
 * Parameter values are updated
 * The negative log-likelihood is calculated
 * The gradient functions return the gradients (vector of 1st derivatives) based on the new parameter values
5. Model convergence is reached when:
 * The gradients are near zero
 * The negative log-likelihood is at a minimum

---

# TMB Model


.pull-left[
linReg.cpp

```cpp
// Simple linear regression
#include <TMB.hpp>
template <class Type>
  
Type objective_function<Type>::operator()()
{
  DATA_VECTOR(y);
  DATA_MATRIX(X);
  PARAMETER_VECTOR(beta);
  PARAMETER(lnSigma);
  
  Type nll = 0;
  Type sigma = exp(lnSigma);
  int n = y.size();
  vector<Type> mu = X * beta;
  for(int i=0; i<n; i++){
    nll -= dnorm(y(i), mu(i), sigma, true);
  }
  
  Type sig2 = pow(sigma,2);
  REPORT(sig2);
  ADREPORT(sig2);
  
  return nll;
}
```
]

.pull-right[
**C++ preliminaries**
* Lines end in semi-colons
* Everything must be declared
* Type is generic and assigned based on input
* Indexing starts at 0!
* x -= 1: x = x-1
* Math operators similar to R (+,-,/,*) 
* Use pow(x,p) for x^p
* if statements cannot be based on a parameter
]



---
# Data: Importing data from R<br><br>
* Pass data to TMB with these 'macros'
* Note: do not specify the object dimension
<br><br>
```{r echo = FALSE, results = 'asis'}
library(knitr)
df <- as.data.frame(
  rbind(
    c('DATA_VECTOR(x)',	'tmbutils::vector<Type>',	'vector'),
    c('DATA_MATRIX(x)','tmbutils::matrix<Type>',	'matrix'),
    c('DATA_SCALAR(x)',	'Type',	'numeric(1)'),
    c('DATA_INTEGER(x)',	'int',	'integer(1)'),
    c('DATA_FACTOR(x)',	'Eigen::vector<int>',	'factor'),
    c('DATA_ARRAY(x)'	,'tmbutils::array<Type>',	'array')
  )
)
colnames(df) <- c('TMB Syntax' , 'C++ Type' , 'R Type')
kable(df)
```
---
# Data: Importing data from R<br><br>

.pull-left-narrow[
TMB code
```cpp
DATA_VECTOR(y); 
DATA_MATRIX(X); 
DATA_INTEGER(i); 
DATA_FACTOR(ngroup);  

```
]

.pull-right-wide[
R script
```{r}
Data <- list(
  y = c(30.2, 45.3, 12.1),
  X = matrix(0,3,3),
  i = 11,
  ngroup = c(1,1,2,2,2)
)
str(Data)
```
]



---
# Declaring model parameters<br><br>
* No PARAMETER_INTEGER
* Again, do not specify the object dimension
<br><br>
```{r echo = FALSE, results = 'asis'}
library(knitr)
df <- as.data.frame(
  rbind(
    c('PARAMETER_VECTOR(x)',	'tmbutils::vector<Type>',	'vector'),
    c('PARAMETER_MATRIX(x)','	tmbutils::matrix<Type>',	'matrix'),
    c('PARAMETER_ARRAY(x)'	,'tmbutils::array<Type>',	'array'),
    c('PARAMETER(x)'	,'Type',	'numeric(1)')
  )
)
colnames(df) <- c('TMB Syntax' , 'C++ Type' , 'R Type')
kable(df)
```

---
# Declaring model parameters<br><br>

.pull-left-narrow[
TMB code
```cpp
PARAMETER_VECTOR(beta); 
PARAMETER(ln_sigma); 
PARAMETER_MATRIX(u); 
```
]

.pull-right-wide[
R script
```{r}
Pars <- list(
   beta = c(0,0),
   lnSigma = 0,
   u = matrix(0,3,3) 
)

str(Pars) 
```
]

