---
title:  "Parameters and Simulation"
subtitle: "TMB Training Session I"
author: "Andrea Havron<br>NOAA Fisheries, OST"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css", "slides-style.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---

layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


<style type="text/css">

code.cpp{
  font-size: 14px;
}
code.r{
  font-size: 14px;
}


</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()
```

---
class: middle

#Dealing with parameters
---
#Parameter Transformations
.pull-left[
Map parameters from real to parameter space
  * Minimizers search for MLE from $-\infty$ to $+\infty$
  * Certain parameters have restricted support:
      * $\sigma^{2} > 0$
        ```cpp
        sigma = exp(ln_sigma)
        ```
      * $0 < \pi < 1$ 
      ```cpp
      pi = 1/(1 + exp(-logit_pi)) # "expit"
      ```
]

.pull-right[
```{r, echo = FALSE, fig.height = 3.5 }
par(cex.lab = 1.5)
fun <- function(x) exp(x)
curve(fun(x), -5, 5, ylab = "sigma", xlab = "")

fun <- function(x) 1/(1 + exp(-x))
curve(fun(x), -5, 5, ylab = "pi", xlab = "")
```
]
---
#Expected Value Transformations

<br>
  * Generalized Linear or Generalized Linear Mixed Models: model a function of the mean that is linear in X:
    $$\text{GLM: }\eta = X\beta$$ 
    $$\text{GLMM: }\eta = X\beta + u$$
    $$E[Y] = g^{-1}(\eta)$$
  * Distributions from the [**exponential family**](https://en.wikipedia.org/wiki/Exponential_family) have natural transformations and inverse link functions:
    $$y \sim \text{Pois}(exp(\eta))$$
    $$y \sim \text{Binomial}(N, expit(\eta))$$

---
#Expected Value Transformations
.three-column[
**Multinomial:** 
$$p_{1}, ... p_{k}$$
$$p_{k} = 1 - \sum^{k-1}_{i=1}p_{i}$$
]  
.three-column[
Natural link function
\begin{bmatrix}
log\frac{p_{1}}{1 - \sum^{k-1}_{i=1}p_{i}}\\
\vdots\\
log\frac{p_{k-1}}{1 - \sum^{k-1}_{i=1}p_{i}}\\
0
\end{bmatrix}

]  
.three-column[
Inverse Link Function
\begin{bmatrix}
\frac{exp(\eta_{1})}{1 + \sum^{k-1}_{i=1}exp(\eta_{i})}\\
\vdots\\
\frac{exp(\eta_{k-1})}{1 + \sum^{k-1}_{i=1}exp(\eta_{i})}\\
\frac{1}{1 + \sum^{k-1}_{i=1}exp(\eta_{i})}
\end{bmatrix}
]  

<br>

$\hspace{1.7in}$**Gamma:**
\begin{align}
E[Y] &= \alpha\beta = exp(\eta)\\
Var[Y] &= \alpha\beta^{2}\\
cv^{2} &= \frac{\alpha\beta^{2}}{\alpha^{2}\beta^{2}} = \frac{1}{\alpha}\\
y &\sim \text{Gamma}(\frac{1}{cv^{2}}, exp(\eta)*cv^{2})
\end{align}
---
#Non-linear Regression

[**Orange tree example**](https://github.com/kaskr/adcomp/blob/master/tmb_examples/orange_big.cpp): Expected size of tree, $i$ at time $t$ is a logistic function of age:<br>
$$\mu_{i,t} = \frac{a}{1 + exp(-(age_{i,t} - b)/c)}$$<br>
a = upper maximum size<br>
b = age when circumference = 1/2a<br>
c = inverse growth rate<br>

$$y_{i,t} \sim N(\mu_{i,t}, \sigma)$$




---
#Parameter mapping

#### TMB allows users to collect and fix parameters using the map argument in MakeADFun()
.large[.p[
* The parameter map is a list of parameters that are fixed or collected in a model run
* Names and dimensions of parameters in the map list must match those of the parameter list
* The map list is structured as a list of factors
* Parameters with factor(NA) are fixed
* Parameters with equal factor levels are collected to a common value
]]

---
#Likelihood Ratio Tests

.pull-left[
$H_{0}$: Null Hypothesis 
.p[
- $\theta \in \Theta_{0}$
- parameter $\theta$ is in a subset of a given parameter space, $\Theta_{0}$
- restricted model
]]
.pull-right[
$H_{1}$: Alternative Hypothesis

.p[
- $\theta \in \Theta$
- parameter $\theta$ is in the complete parameter space, $\Theta$
- unrestricted, full model
]]

<br>
**Likelihood Ratio Test**:
$$LR = -2[\ell(\theta_{0}) - \ell(\hat{\theta})]$$

$$LR \sim \chi^{2}(df_{\chi^{2}} = df_{\Theta}-df_{\Theta_{0}})$$

---
# Parameter mapping with LR test
<br>
TMB example: [**lr_test.cpp**](https://github.com/kaskr/adcomp/blob/master/tmb_examples/lr_test.cpp)

```cpp
/ Illustrate map feature of TMB to perform likelihood ratio tests on a ragged array dataset.
#include <TMB.hpp>

template<class Type>
Type objective_function<Type>::operator() ()
{
  DATA_VECTOR(obs);
  DATA_FACTOR(group);
  
  PARAMETER_VECTOR(mu);
  PARAMETER_VECTOR(sd);
  
  Type res=0;
  
  for(int i=0;i<obs.size();i++){
    res -= dnorm(obs[i],mu[group[i]],sd[group[i]],true);
  }
  
  return res;
}

```
---

class: middle

# Simulation
---
# Simulation in TMB

* Standard generator functions to simulate data within the TMB model:

```cpp
rnorm()           rpois()           runif()
rbinom()          rgamma()          rexp()          
rbeta()           rf()              rlogis()  
rt()              rweibull()        rcompois()
rtweedie()        rnbinom()         rnbinom2()
```


* Simulation blocks are used to call simulations from R

.pull-left[
**Binomial Example**<br>
C++ code
```cpp
for(int i = 0; i < n_obs; i++){
  eta(i) = beta(0) + beta(1)*x(i); 
  p(i) = 1/(1 + exp(-eta(i)));
  nll(i) = -dbinom(Y(i), N(i), p(i), true); 
  SIMULATE{
    Y(i) = rbinom(N(i), p(i));
  }
}
SIMULATE{
  REPORT(Y);
}
```
]

.pull-right[
<br>
R code
```{r, eval = FALSE}
set.seed(1) ## optional

obj$simulate(complete = TRUE)
```
*Note*
* optimization not necessary for simulation
* not vectorized like in R
]

---
#Simulation Study

.pull-left[
linReg.cpp
```cpp
 for(int i=0; i<n; i++){
    nll -= dnorm(y(i), mu(i), sigma, true);
  }
  SIMULATE{
    y = rnorm(mu, sigma);
    REPORT(y);
  }
```
Set-up Model in R
```{r}
library(TMB)
dyn.load(dynlib('../src/linReg'))
sig <- 2
beta <- c(0.5,2)

Data <- list(y = rep(0,50), 
             X =  cbind(rep(1,50), 1:50))
Pars <- list(beta = c(0.5,2),lnSigma = log(2))
obj <- MakeADFun(data = Data, 
                 parameters = Pars, 
                 DLL = 'linReg')
obj$par
```
]
.pull-right[
Run Simulation
```{r}
set.seed(1)
sim <- replicate(500, {
  simdata <- obj$simulate(par=obj$par, complete=TRUE)
  obj2 <- MakeADFun(simdata, Pars, 
                    DLL="linReg", silent=TRUE)
  nlminb(obj2$par, obj2$fn, obj2$gr)$par
})
obj$par
apply(t(sim), 2, mean)
```
]
