---
title:  "TMB Checking Output and Validation"
subtitle: "FIMS Implementation Team Workshop"
author: "Andrea Havron<br>NOAA Fisheries, OST<br> 2022-03-23"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css", "slides-style.css"]
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


<style type="text/css">

code.cpp{
  font-size: 14px;
}
code.r{
  font-size: 14px;
}


</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()
```

---
# Debugging: Error types<br><br>
.pull-left[
1. Compile time errors
2. TMB-R linkage errors
3. Runtime errors
4. Convergence errors
5. Validation errors
]
---

# Debugging: Error types<br>
.pull-left[
1. **Compile time errors**
2. TMB-R linkage errors
3. Runtime errors
4. Convergence errors
5. Validation errors
]

.pull-right[
Fails with errors when compiling
* Incorrect syntax: e.g., missing **)** or **;**
* Wrong type, for example:
  * int instead of Type
  * passing matrix to a function that accepts a vector
* Undeclared variable
]

.pull-left[
linReg.cpp
```cpp
DATA_VECTOR(y
DATA_MATRIX(X);

PARAMETER_VECTOR(beta);

int nll;
matrix<Type> mu = X*beta;
nll = -dnorm(y,mu,sigma,true)
return nll;
```
]


---
# Debugging: Error types<br>
.pull-left[
1. **Compile time errors**
2. TMB-R linkage errors
3. Runtime errors
4. Convergence errors
5. Validation errors
]

.pull-right[
```cpp
DATA_VECTOR(y
DATA_MATRIX(X);

PARAMETER_VECTOR(beta);

int nll;
matrix<Type> mu = X*beta;
nll = -dnorm(y,mu,sigma,true)
return nll;
```
]

.pull-left[
linReg.R
```{r, eval = FALSE}
setwd("2022_TMB_Session_I")

#compilation flags when using gdbsource

#On Linus/OS X
TMB::compile("src/debug1.cpp","-O0 -g")

#On Windows
TMB::compile("src/debug1.cpp","-O1 -g",
             DLLFLAGS="")
```
]

.pull-right[
Console
```{r, eval = FALSE}
#Debug using RStudio:
> TMB:::setupRStudio()
> TMB::compile("src/debug1.cpp")

#Debug using gdbsource
#hangs on Windows
> TMB::gdbsource("R/debug_gdbsource.R")
#works on Windows but not always 
#informative
> TMB::gdbsource("R/debug_gdbsource.R", 
                 interactive = TRUE)
```
]


---

# Debugging: Error types<br>
.pull-left[
1. Compile time errors
2. **TMB-R linkage errors**
3. Runtime errors
4. Convergence errors
5. Validation errors
]

.pull-right[
Compiles but MakeADFun crashes
* .dll not loaded in R
* Incorrect data/parameter types or missing value
* Dimension mismatch
]

.pull-left[
```cpp
DATA_VECTOR(y);
DATA_MATRIX(X);

PARAMETER_VECTOR(beta);
PARAMETER(lnSigma);
Type sigma = exp(lnSigma);

Type nll = 0;
vector<Type> mu = X*beta;
nll -= sum(dnorm(y,mu,sigma,true));
return nll;
```
]

.pull-right[
linReg.R
```{r, eval = FALSE}
library(TMB)
compile(linReg.cpp)
Data <- list(y = c(-2.1, 3.3, 4.2),
             X = data.frame(c(1,1),c(4,5)))
Pars <- list(beta = 0)
obj <- MakeADFun(data = Data, 
                 parameters = Pars, 
                 DLL = "linReg")
```
]

---

# Debugging: Error types<br>
.pull-left[
1. Compile time errors
2. TMB-R linkage errors
3. **Runtime errors**
4. Convergence errors
5. Validation errors
]

.pull-right[
Model fails during minimization
  * Data-distribution mismatich 
  * Incorrect parameter transformation
  * Not taking log of the likelihood
  * Illegal mathematical operations, eg. log(-1)
]

.pull-left[
debug2.cpp
```cpp
DATA_VECTOR(y);
DATA_MATRIX(X);

PARAMETER_VECTOR(beta);
PARAMETER(lnSigma);

Type nll = 0;
Type sigma = exp(lnSigma);
vector<Type> mu = X*beta;
nll -= sum(dlnorm(y,mu,sigma));
return nll;
```
]

.pull-right[
simple.R
```{r, eval = FALSE}
library(TMB)
compile("debug2.cpp")
dyn.lib(dynload("debug2"))
Data <- list(y = c(-2.1, 3.3, 4.2),
             X = cbind(c(1,1,1),c(4,5,2)))
Pars <- list(beta = c(0,0), sigma = 0)
obj <- MakeADFun(data = Data, 
                 parameters = Pars, 
                 DLL = "linReg")
opt <- nlminb(obj$par, obj$fn, obj$gr)
```
]

---

# Debugging: Error types<br>
.pull-left[
1. Compile time errors
2. TMB-R linkage errors
3. Runtime errors
4. **Convergence errors**
5. Validation errors
]

.pull-right[
 Model compiles and runs but fails to converge
  * Singular convergence: model is likely overparameterized
  * False convergence: likelihood may be discontinuous
  * Relative convergence but Hessian not positive definite: singularity
]

**Non-Invertible Hessians**
* The Hessian will **not be invertible** if the MLE is not a true minimum
* When could this occur? Usually mis-specified models
* Parameters confounded or overparameterized (too complex for data)
* TMB will warn about uninvertible Hessians (NaNs)

---

# Debugging: Error types<br>
.pull-left[
1. Compile time errors
2. TMB-R linkage errors
3. Runtime errors
4. Convergence errors
5. **Validation errors**
]

.pull-right[
 Model compiles and runs and passes convergence tests but is incorrect
  * Model structure is mis-specified
  * Incorrect distribution for observations
  * Incorrect random effect structure
]




