---
title:  "Modular TMB Example"
subtitle: "TMB Training Session I"
author: "Andrea Havron<br>NOAA Fisheries, OST"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css", "slides-style.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---

layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


<style type="text/css">

code.cpp{
  font-size: 14px;
}
code.r{
  font-size: 14px;
}


</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()
```

---
# C++ Anatomy of a Class

```cpp
class ClassName{
  Access specifier: //can be private, public, or protected
  Data members; //variables to be used
  
  ClassName(){} //Constructor - automatically called when new object created
  
  Member functions(){} //Methods to access data members
  
}; //class name ends with semicolon


//create instance of class
ClassName<T>* ClassName<T>::instance = new ClassName<T>();

```
.p[
- Memory is not allocated when a class is defined
- An instance of the class needs to be created for memory allocation (new object created)
- **new** is the C++ initialize function that allocates memory
- Once the class is instantiated, members can be accessed using the dot('.')
]
resources:<br>
[**geeksforgeeks**](https://www.geeksforgeeks.org/c-classes-and-objects/)<br>
[**cplusplus**](https://cplusplus.com/doc/tutorial/classes/)
---
# FIMS Dmultinom class
[inst/include/distributions/functors/tmb_distributions.hpp](https://github.com/NOAA-FIMS/FIMS/blob/main/inst/include/distributions/functors/tmb_distributions.hpp)

.pull-left-wide[
```cpp
#ifdef TMB_MODEL

template <typename T>
struct Dmultinom : public DistributionsBase<T> {
  /** EigenVector defined in interface.hpp */
  using Vector = typename ModelTraits<T>::EigenVector;
  Vector x; /*!< Vector of length K of integers */
  Vector p; /*!< Vector of length K, specifying the probability for the K
               classes (note, unlike in R these must sum to 1). */

  Dmultinom() : DistributionsBase<T>() {}

  /**
   * @brief Probability mass function of the multinomial distribution.
   *
   * @param do_log Boolean; if true, log densities are returned
   */
  virtual const T evaluate(const bool& do_log) {
    return dmultinom(x, p, do_log);
  }
};

#endif

```
]
.pull-right-narrow[
Access class members and functions
<br>
```cpp
fims::Dmultinom<T> ll_dmultinom;
ll_dmultinom.x = x;
ll_dmultinom.p = p;
ll_dmultinom.evaluate(true);
```
]

---
class: middle
.large[
[**Model used to test Dmultinom**](https://github.com/NOAA-FIMS/FIMS/blob/193-feature-rcpp-interface-to-call-module-likelihoods/inst/extdata/TMB_tests/distributions/test_dmultinom_distribution.hpp)


[**Corresponding TMB Model**](https://github.com/NOAA-FIMS/FIMS/blob/193-feature-rcpp-interface-to-call-module-likelihoods/inst/extdata/TMB_tests/distributions/test_dmultinom_distribution.cpp)
]