---
title:  "TMB 101 Exercises"
subtitle: "FIMS Implementation Team Workshop"
author: "Andrea Havron<br>NOAA Fisheries, OST<br> 2022-03-23"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css", "slides-style.css"]
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


<style type="text/css">

code.cpp{
  font-size: 14px;
}
code.r{
  font-size: 14px;
}


</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()
```

---
#Maturity at age model<br>
The number of mature fish, $C_{y,a}$ out of the total number of fish, $N_{y,a}$ at year, $y$ and age, $a$, can be expressed as a binomial distribution,

$$C_{y,a} \sim Binom(N_{y,a}, p^{M}_{y,a})$$,

where $p^{M}$ represents the probability of maturity, which can be expressed as a logistic transformation of a slope parameter and $a_{50}$, the age of 50% maturity,

\begin{align}
\beta &= c(-slope*a50, slope)\\
\eta_{y,a} &= \beta_{0} + \beta_{1}*a\\
p^{M}_{y,a} &= \frac{1}{1+exp(-\eta_{y,a})}
\end{align}

---
#Maturity at age model<br>

.pull-left[
```cpp
#include <TMB.hpp>

template<class Type>
Type objective_function<Type>::operator()()
{
  DATA_VECTOR(Y);
  DATA_VECTOR(N);
  DATA_VECTOR(age_obs); 
  DATA_INTEGER(max_age); 
  PARAMETER_VECTOR(beta); 

  int n_obs = Y.size();
  vector<Type> nll(n_obs);
  nll.setZero();
     
  vector<Type> logit_mat(n_obs);
  vector<Type> mat(n_obs);
  for(int i = 0; i < n_obs; i++) {
    logit_mat(i) = beta(0) + 
       beta(1)*age_obs(i); 
    mat(i) = 1/(1 + exp(-logit_mat(i)));
    nll(i) = -dbinom(Y(i), N(i), mat(i),1); 
  }

  return sum(nll);
}
```
]

.pull-right[
Simulate data
```{r, eval = FALSE}
a50 = 5
slope = 2
beta = c(-slope*a50,slope)
ages = t(matrix(1:20, 20, 40))
mat = 1/(1+exp(-(beta[1] + beta[2]*ages)))
set.seed(123)
samp <- sample(50:100, length(ages),
               replace=TRUE)
N = matrix(samp, nrow = 20)
ysim <- rbinom(length(N), N, mat)
Y = matrix(ysim, nrow = 20)
```
]
---

#Maturity at age model<br>

.pull-left[
```cpp
#include <TMB.hpp>

template<class Type>
Type objective_function<Type>::operator()()
{
  DATA_VECTOR(Y);
  DATA_VECTOR(N);
  DATA_VECTOR(age_obs); 
  DATA_INTEGER(max_age); 
  PARAMETER_VECTOR(beta); 

  int n_obs = Y.size();
  vector<Type> nll(n_obs);
  nll.setZero();
     
  vector<Type> logit_mat(n_obs);
  vector<Type> mat(n_obs);
  for(int i = 0; i < n_obs; i++) {
    logit_mat(i) = beta(0) + 
       beta(1)*age_obs(i); 
    mat(i) = 1/(1 + exp(-logit_mat(i)));
    nll(i) = -dbinom(Y(i), N(i), mat(i),1); 
  }

  return sum(nll);
}
```
]

.pull-right[
Run model
```{r, eval = FALSE}
library(TMB)
compile("maturity_0.cpp")
dyn.load(dynlib("maturity_0"))

input = list(data=list(),par=list())
input$par$beta = c(0,0)
input$data$Y = c(Y)
input$data$N = c(N)
input$data$age_obs = c(ages)
input$data$max_age = max(ages)

mod = MakeADFun(input$data, input$par, 
                DLL = "maturity_0")

opt = nlminb(mod$par, mod$fn, mod$gr)
```
]

