---
title:  "TMB Checking Output and Validation"
subtitle: "FIMS Implementation Team Workshop"
author: "Andrea Havron<br>NOAA Fisheries, OST<br> 2022-03-23"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css", "slides-style.css"]
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


<style type="text/css">

code.cpp{
  font-size: 14px;
}
code.r{
  font-size: 14px;
}


</style>


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()
```

---
# Debugging: Error types<br><br>
.pull-left[
1. Compile time errors
2. TMB-R linkage errors
3. Runtime errors
4. Convergence errors
5. Validation errors
]
---
#Maturity at age model<br>
The number of mature fish, $Y_{y,a}$ out of the total number of fish, $N_{y,a}$ at year, $y$ and age, $a$, can be expressed as a binomial distribution,

$$Y_{y,a} \sim Binom(N_{y,a}, p_{y,a})$$,

where $p$ represents the probability of maturity. Common to estimate maturity using logistic regression: 

$$log\left(\frac{p_{y,a}}{1-p_{y,a}}\right) = \eta_{y,a} = \beta_0 + \beta_1 a$$

The age of 50% maturity is $a_{50} = -\frac{\beta_0}{\beta_1}$,

---

#Overdispersed model<br>
.pull-left[
```cpp
#include <TMB.hpp>
template<class Type>
Type dbetabinom(Type x, Type n, Type p, Type phi, int do_log)
{
  Type ll = lgamma(n + 1.0) - lgamma(x + 1.0) - lgamma(n - x + 1.0) + 
    lgamma(x + p*phi) + lgamma(n - x +(1-p)*phi) - lgamma(n + phi) +
    lgamma(phi) - lgamma(p*phi) - lgamma((1-p)*phi);
  if(do_log) return(ll);
  else {
    return(exp(ll));  
  }
}
```
See 
* [maturity_bb.R](https://github.com/NOAA-FIMS/TMB_training/blob/main/maturity_examples/maturity_bb.R)
* [maturity_bb.cpp](https://github.com/NOAA-FIMS/TMB_training/blob/main/maturity_examples/maturity_bb.cpp)
]

.pull-right[
```cpp
template<class Type>
Type objective_function<Type>::operator()()
{
  DATA_VECTOR(Y); //number mature
  DATA_VECTOR(N); //number of mature + not mature
  DATA_VECTOR(age_obs); //age for each observation
  DATA_INTEGER(max_age); //to generate a maturity ogive
  PARAMETER_VECTOR(beta); //intercept, slope
  PARAMETER(log_phi);
  Type phi = exp(log_phi);

  int n_obs = Y.size();
  vector<Type> nll(n_obs);
  nll.setZero();
     
  vector<Type> logit_mat(n_obs), mat(n_obs);
  for(int i = 0; i < n_obs; i++) {
    logit_mat(i) = beta(0)  + beta(1)*age_obs(i); 
    mat(i) = 1/(1 + exp(-logit_mat(i)));
    nll(i) = -dbetabinom(Y(i), N(i), mat(i), phi, 1); //negative log-likelihood
  }

  return sum(nll);
}
```
]

---

# Debugging: Error types<br>
.pull-left[
1. **Compile time errors**
2. TMB-R linkage errors
3. Runtime errors
4. Convergence errors
5. Validation errors
]

.pull-right[
Fails with errors when compiling
* Incorrect syntax: e.g., missing **)** or **;**
* Wrong type, for example:
  * int instead of Type
  * passing matrix to a function that accepts a vector
* Undeclared variable
]

.pull-left[
linReg.cpp
```cpp
DATA_VECTOR(y
DATA_MATRIX(X);

PARAMETER_VECTOR(beta);

int nll;
matrix<Type> mu = X*beta;
nll = -dnorm(y,mu,sigma,true)
return nll;
```
]


---
# Debugging: Error types<br>
.pull-left[
1. **Compile time errors**
2. TMB-R linkage errors
3. Runtime errors
4. Convergence errors
5. Validation errors
]

.pull-right[
```cpp
DATA_VECTOR(y
DATA_MATRIX(X);

PARAMETER_VECTOR(beta);

int nll;
matrix<Type> mu = X*beta;
nll = -dnorm(y,mu,sigma,true)
return nll;
```
]

.pull-left[
linReg.R
```{r, eval = FALSE}
#compilation flags when using gdbsource

#On Linus/OS X
TMB::compile("src/linReg.cpp","-O0 -g")

#On Windows
TMB::compile("src/linReg.cpp","-O1 -g",
             DLLFLAGS="")
```
]

.pull-right[
Console
```{r, eval = FALSE}
#Debug using RStudio:
> TMB:::setupRStudio()
> TMB::compile("src/linReg.cpp")

#Debug using gdbsource
#hangs on Windows
> TMB::gdbsource("R/linReg.R")
#works on Windows but not always 
#informative
> TMB::gdbsource("R/linReg.R", 
                 interactive = TRUE)
```
]


---

# Debugging: Error types<br>
.pull-left[
1. Compile time errors
2. **TMB-R linkage errors**
3. Runtime errors
4. Convergence errors
5. Validation errors
]

.pull-right[
Compiles but MakeADFun crashes
* .dll not loaded in R
* Incorrect data/parameter types or missing value
* Dimension mismatch
]

.pull-left[
```cpp
DATA_VECTOR(y);
DATA_MATRIX(X);

PARAMETER_VECTOR(beta);
PARAMETER(lnSigma);
Type sigma = exp(lnSigma);

Type nll = 0;
vector<Type> mu = X*beta;
nll -= sum(dnorm(y,mu,sigma,true));
return nll;
```
]

.pull-right[
linReg.R
```{r, eval = FALSE}
library(TMB)
compile(linReg.cpp)
Data <- list(y = c(-2.1, 3.3, 4.2),
             X = data.frame(c(1,1),c(4,5))
Pars <- list(beta = 0)
obj <- MakeADFun(data = Data, 
                 parameters = Pars, 
                 DLL = "linReg")
```
]

---

# Debugging: Error types<br>
.pull-left[
1. Compile time errors
2. TMB-R linkage errors
3. **Runtime errors**
4. Convergence errors
5. Validation errors
]

.pull-right[
Model fails during minimization
  * Data-distribution mismatich 
  * Incorrect parameter transformation
  * Not taking log of the likelihood
  * Illegal mathematical operations, eg. log(-1)
]

.pull-left[
linReg.cpp
```cpp
DATA_VECTOR(y);
DATA_MATRIX(X);

PARAMETER_VECTOR(beta);
PARAMETER(sigma);

Type nll = 0;
vector<Type> mu = log(X*beta);
nll -= sum(dlnorm(y,mu,sigma));
return nll;
```
]

.pull-right[
simple.R
```{r, eval = FALSE}
library(TMB)
compile(linReg.cpp)
dyn.lib(dynload("linReg"))
Data <- list(y = c(-2.1, 3.3, 4.2),
             X = cbind(c(1,1,1),c(4,5,2))
Pars <- list(beta = c(0,0), sigma = 0)
obj <- MakeADFun(data = Data, 
                 parameters = Pars, 
                 DLL = "linReg")
opt <- nlminb(obj$par, obj$fn, obj$gr)
```
]

---

# Debugging: Error types<br>
.pull-left[
1. Compile time errors
2. TMB-R linkage errors
3. Runtime errors
4. **Convergence errors**
5. Validation errors
]

.pull-right[
 Model compiles and runs but fails to converge
  * Singular convergence: model is likely overparameterized
  * False convergence: likelihood may be discontinuous
  * Relative convergence but Hessian not positive definite: singularity
]

**Non-Invertible Hessians**
* The Hessian will **not be invertible** if the MLE is not a true minimum
* When could this occur? Usually mis-specified models
 * Parameters confounded or overparameterized (too complex for data)
* TMB will warn about uninvertible Hessians (NaNs)

---

# Debugging: Error types<br>
.pull-left[
1. Compile time errors
2. TMB-R linkage errors
3. Runtime errors
4. Convergence errors
5. **Validation errors**
]

.pull-right[
 Model compiles and runs and passes convergence tests but is incorrect
  * Model structure is mis-specified
  * Incorrect distribution for observations
  * Incorrect random effect structure
]

---
# Implementing OSA in TMB<br>

.pull-left[
```cpp
 #include <TMB.hpp>

template<class Type>
Type objective_function<Type>::operator() ()
{
  DATA_VECTOR(y);         
  DATA_SPARSE_MATRIX(B);  
  DATA_SPARSE_MATRIX(X); 
  DATA_VECTOR_INDICATOR(keep, y); 
  PARAMETER_VECTOR(u);    
  PARAMETER_VECTOR(beta); 
  PARAMETER(lnSDu);      
  PARAMETER(lnSDy);    

...

  Type nll = 0;
  Type sdu = exp(lnSDu);
  nll -= dnorm(u, Type(0), sdu, true).sum();

...

   vector<Type> mu = X * beta + B * u;
  Type sdy = exp(lnSDy);
  for(int i=0; i<y.size(); i++){
    nll -= keep(i) * dnorm(y(i), mu(i), sdy, true);
  }
  
  ...
  
  return nll;
}


```
]


* See [src/simple_re.cpp](https://github.com/NOAA-FIMS/TMB_training/blob/main/src/simple_re.cpp)
* [maturity_bb_diagnostics.R](https://github.com/NOAA-FIMS/TMB_training/blob/main/maturity_examples/maturity_bb_diagnostics.R)
]


